Time for a rethink: time sub-sampling methods in disparity-through-time analyses

### 2018-02-06 manuscript/response_to_reviewers.tex

We realise we had not made this clear, thus we have added the following explanation in the text:

\textit{For example, using the ``equal.splits'' model on an ancestor and a descendant with PCO1 values of respectively $0.04$ and $0.03$, after a sufficient number of bootstrap replicates (e.g. 100) the value along the branch will be close to $0.5\times0.04 + 0.5\times0.03 = 0.035$.
By estimating this value rather than generating it (i.e. creating a new element mid-way along the branch that would be the average of the descendant and ancestor - $0.035$) we obtain the same results without modifying the morphospace properties.} lines @@@

%TG: I can implement that pretty easily but I think that mathematically it's not cool (we can not add values post-ordination in the ordinated space). Two options: (1) implementing the suggestion but with heavy warnings; (2) writing down a clever response to this comment. In the dispRity package review, that was highlighted (I bet reviewer 2 is Graeme) and I responded something similar to option (2) so I'd rather go that way for consistency.

% NC: I like this response. I dunno if it is Graeme, they use US spellings... Maybe Bapst?

@@

\begin{enumerate}

\item{\textcolor{blue}{It is not immediately clear what is meant by ``defined by the average duration of the stratigraphic period'' (line 337, 341) vs ``defined by the average number of stratigraphic periods'' (line 348, 353). After thinking about it for awhile and then checking the R code, this is what appears to be going on. Let's say we have a case study that spans 10 stratigraphic intervals over 40 million years. Thus for the stratigraphy approach, there are 10 sub-samples which vary in their duration. Let’s say the average duration of those intervals is 5 million years. For the duration approach, the 40 million year span is divided into 5-million-year bins, resulting in 8 sub-samples. For the number approach, the 40 million year span is divided into 10, each of which is 4 million years long. If this is the case, the use of the word ``average'' when describing the number approach is confusing. I am also having trouble reconciling this with the number of points in each panel in Figure 2, unless there is also a minimum sample size applied (for example in the Bapst 2016 study, there are only three disparity values for the time-binned series using the ``number'' approach, maybe because after dividing the time scale up this way, there were not enough taxa in the first and second time bins?). It may be helpful to include a graphic, even if in relegated to one of the appendices.}}

%TG: good idea, we (I) can make a figure designating the three different strategies.

\item{\textcolor{blue}{Line 158. Gower's distance is not the Euclidean distance (= the square root of the squared distances) over the number of shared characters. For a set of unordered characters, it is the number of mismatched character states over the number of comparable/shared characters. For a set of ordered characters, it is sum of the differences in the values of the characters states for each pairwise comparison, over the number of comparable characters. This is an edit that needs to be made only to the text; because the authors used functions from the Claddis package (see github documentation), they would have calculated Gower's distance correctly.}}

@@

\item{\textcolor{blue}{Could I suggest adding ‘sub-sampling’ to your key words? We now recommend that as many key words as possible are also included in your title, as well as being well-used in the abstract.}}

Done.

\item{\textcolor{blue}{Please upload your supporting information files to Dryad as separate files (with the table in an editable format).}}

We have split the appendices into three files S1) additional details of the datasets; S2) additional figures; and S3) additional tables (now in .docx format rather than PDF).
% NC: Just need to sort S2 when we have the new figure. I've added them as is to Dryad to get the reference link.

@@

\item{\textcolor{blue}{Where you have two references with the same citation (e.g. Bapst et al. 2016), please list them in the same order as they are cited in the text, irrespective of the second author name (currently Bapst et al. 2016b and Brusatte et al. 2014b are both cited before their respective a references; in each case the first reference cited should be a and be listed first).}}

Sorry this is a bug in our reference manager software. Now fixed!

\item{\textcolor{blue}{Guillerme (2016) reference requires some more information – is there a URL for the software that can be added? (Similar for any other software reference, e.g. Lloyd 2015.)}}

We have fixed this.
% NC: Fix at converting to Word stage, hopefully there will be a paper to cite by then... for now I've used the v 0.5 zenodo citation. I've changed the citation of Claddis to Graeme's Biol Linn soc paper.

### 2018-02-07 manuscript/response_to_reviewers.tex

We apologise for the confusion in the description of Figure 3 and have fixed it (see Reviewer 1, comment \ref{fig3_feckup}). We were rushing to get this submitted by the symposium deadline and this error slipped through the cracks.

\item{\textcolor{blue}{Finally, I have spent some time thinking about by the statement ``all time binning approaches (whether equally sized or not) assume that characters of taxa evolve following a punctuated equilibrium model.'' (lines 70-71). I see what the authors are getting at — they are worried that doing this requires an assumption of constant evolution across the time bin. But I would argue that it is not necessary to interpret it this way. Instead it could simply represent an average across the interval sampled without the need to make any assumption about how change occurred within that interval. Plus, even in scenarios where there was enough change that at a course scale, it appeared that disparity was static, some of that volatility might be revealed if the time-scale were altered.}}

We've tone down this statement by included the fact that bins can simply represent averages (as suggested by the reviewer).

\textit{Second, the time binning approaches (whether bins are equally sized or not) favour punctuated equilibrium modes of evolution.
Whether the disparity represents an average across the interval (with no interpretation on if or how it varied within the time bin) or it is effectively postulated to be constant, when analysing the changes in disparity through time, this method will only allow changes in disparity (in average or constant) to occur only between intervals rather than also allowing for gradual changes within intervals (a pattern that is fairly common in the fossil record, CITATION).} lines @@@

%TG: tone done the statement? % NC: yeah, not sure what else we can do. %TG: Does that kinda work?

@@

\item{\textcolor{blue}{But there is also a deeper issue here, in that the fossil record is at its heart a time-binned record: rates of sedimentation vary and sediment deposition is sporadic, such that local beds represent different amounts of time, or different ``packages'' of time, separated by gaps, and the fossils within them may be time-averaged to various degrees. [...] Therefore, even if you pick a ``point in time'' and count the number of branches/ranges that cross that point, some of those are being captured because the resolution of the end points only allows us to know that the taxon started and/or ended somewhere in the interval encompassing that point in time.}}

%TG: this is a fair point we should include in the discussion.
%TG: though that is just a "practical" point for the vertebrate fossil record. Foraminiferas, for example, do not behave this way (or at a way finer scale).

We've add the following caveat to the discussion:

\textit{Finally, the difference between the time-binning and time-slicing methods could also simply be favoured towards time binning due to the nature of the fossil record itself.
The rates of sedimentation varies in time a space and can be really sporadic, influencing the groups found within and their temporal distribution (e.g. vertebrates, c.f. foraminifera).
In this case, different beds could represent different ``packages'' of fauna through time separated by gaps thus effectively binning the data rather than slicing.
For example, slicing through such strata will yield similar results whether where in time does the slice occur.
It is important to note however, that, as specified above, the time slicing method also includes ancestral estimations data (whether through the nodes or the branches) that are by definition not available in the fossil record and thus are not influenced by its nature.} lines @@@

### 2018-02-07 manuscript/time-slice.tex
respond to couple of reviewer 2 major comments
also changes time-slices (non equal) to time-slices
(equidistant/nonequidistant).

First, time bins defined by stratigraphy are not of equal size, biasing higher disparity towards longer stratigraphic periods. 
This can be dealt with using rarefaction methods, i.e. repeating the analysis while resampling the taxa to have the same number of taxa in each bin (e.g. using bootstrapping with limited resampling).
This can, however, lead to large confidence intervals when there are stratigraphic periods with few species.
Other studies split large time bins so they are of roughly equal size, but this is often an \textit{ad hoc} procedure that can introduce more bias depending on where bins are split.
Second, all time binning approaches (whether bins are equally sized or not) assume that characters of taxa evolve following a punctuated equilibrium model, because disparity is only estimated once for each interval resulting in all changes in disparity occurring \textit{between} intervals, rather than also allowing for gradual changes \textit{within} intervals \citep[a pattern that is fairly common in the fossil record;][]{Hunt21042015}.
Third, when investigating changes in disparity due to events at a specific time point (e.g. a mass extinction), time bins may have not have high enough resolution to resolve changes at the event, for example if time bins are every 20 million years it may be hard to capture the effects of an event five million years into the bin.
Finally, time bin analyses are often limited by the number of taxa in each bin.
If there are insufficient taxa in a time bin, disparity cannot be calculated, so further analyses, e.g. correlations of disparity with hypothesised drivers of morphological evolution, are not possible.

@@

\begin{enumerate}
  \item Stratigraphy: time sub-samples defined by stratigraphic periods. 
  \begin{enumerate}[(i)]
    \item Time bins (unequal). 
    We calculated disparity for the taxa in each stratigraphic period (age or epoch). 
    To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement for each time bin by randomly resampling with replacement a new sub-sample of taxa from the observed taxa in the bin 100 times.
    %TG: actually in the code the data is bootstrapped 100 times only. Should I change to 1000? To be fair, I think 1000 is too much (in terms of calculation and results) and 100 to few. A compromise could be 500.
    %NC: 100 is fine, I was just guessing!
    We then calculated the median disparity value for each time bin along with the 50\% and 95\% confidence intervals.
    \item Time-slices (unequal).
    We calculated disparity using our time-slicing approach with slices occurring at the midpoint of each stratigraphic period (age or epoch), and using all six time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits).
    To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement as described above for the stratigraphic time bins.
    \end{enumerate}

  \item Duration: time sub-samples defined by the duration of stratigraphic periods. 
  \begin{enumerate}[(i)]
    \item Time bins (equal). 
    We calculated disparity for the taxa in each time bin where time bin size was defined by the average \textit{duration} of the stratigraphic period (age or epoch), and bootstrapped the disparity values as described above.
    \item Time-slices (unequal).
    We calculated disparity using our time-slicing approach where the interval between slices, was defined by the average \textit{duration} of the stratigraphic period (age or epoch).
    We used the six time-slicing methods and bootstrapped as described above.
  \end{enumerate}

  \item Number: time sub-samples defined by the number of stratigraphic periods. 
  \begin{enumerate}[(i)]
    \item Time bins (equal). 
    We calculated disparity for the taxa in each time bin where the number of time bins was defined by the average \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest, and bootstrapped the disparity values as described above.
    \item Time-slices (unequal).
    We calculated disparity using our time-slicing approach where the number of slices, was defined by the average \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest.
    We used the six time-slicing methods and bootstrapped as described above.
  \end{enumerate}

@@

Time-slicing has several advantages over time binning (using either equally or unequally sized bins) approaches.
First, it allows us to use as much of the information available to us, in the form of phylogenetic relationships and ancestral taxa, as possible. 
This increases our ability to investigate key biological questions, such as how do various drivers influence morphological diversity through time, and how do mass extinctions influence disparity \citep{Brusatte12092008,Foote29111996,friedmanexplosive2010}, both by increasing the statistical power of analyses and through the availability of data at key time points in the history of our groups.
Second, we are able to be more explicit about the mode of evolution in our clades; in time-slicing we can apply punctuated or gradual models of trait change rather than making an assumption of punctuated evolution.
This may be important, as gradual change is a common pattern of trait evolution in the fossil record \citep{Hunt20112007}.
Of course the method also has limitations.
The main one of these is a practical one; it requires a time-calibrated phylogeny.
Not all palaeontological datasets have phylogenies yet.
Another limitation is that the method, like most phylogeny based methods, depends on ancestral state estimations.
Care should be taken in interpreting these, as they are highly dependent on the data and models used for the estimations \citep{Slateretal2012,doi:10.1080/10635150801910451}.

>>>>

First, time bins defined by stratigraphy are not of equal size, biasing higher disparity towards longer stratigraphic periods. 
This can be dealt with using rarefaction methods, i.e. repeating the analysis while resampling the taxa to have the same number of taxa in each bin (e.g. using bootstrapping with limited resampling).
This can, however, lead to large confidence intervals when there are stratigraphic periods with few species.
Other studies split large time bins so they are of roughly equal size, but this is often an \textit{ad hoc} procedure that can introduce more bias depending on where bins are split.
Second, the time binning approaches (whether bins are equally sized or not) favour punctuated equilibrium modes of evolution.
Whether the disparity represents an average across the interval (with no interpretation on if or how it varied within the time bin) or it is \textit{effectively} postulated to be constant, when analysing the changes in disparity through time, this method will only allow changes in disparity (in average or constant) to occur only \textit{between} intervals rather than also allowing for gradual changes \textit{within} intervals \citep[a pattern that is fairly common in the fossil record;][]{Hunt21042015}.
Third, when investigating changes in disparity due to events at a specific time point (e.g. a mass extinction), time bins may have not have high enough resolution to resolve changes at the event, for example if time bins are every 20 million years it may be hard to capture the effects of an event five million years into the bin.
Finally, time bin analyses are often limited by the number of taxa in each bin.
If there are insufficient taxa in a time bin, disparity cannot be calculated, so further analyses, e.g. correlations of disparity with hypothesised drivers of morphological evolution, are not possible.

@@

\begin{enumerate}
  \item Stratigraphy: time sub-samples defined by stratigraphic periods. 
  \begin{enumerate}[(i)]
    \item Time bins (unequal). 
    We calculated disparity for the taxa in each stratigraphic period (age or epoch). 
    To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement for each time bin by randomly resampling with replacement a new sub-sample of taxa from the observed taxa in the bin 100 times.
    %TG: actually in the code the data is bootstrapped 100 times only. Should I change to 1000? To be fair, I think 1000 is too much (in terms of calculation and results) and 100 to few. A compromise could be 500.
    %NC: 100 is fine, I was just guessing!
    We then calculated the median disparity value for each time bin along with the 50\% and 95\% confidence intervals.
    \item Time-slices (non-equidistant).
    We calculated disparity using our time-slicing approach with slices occurring at the midpoint of each stratigraphic period (age or epoch), and using all six time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits).
    To reduce the influence of outliers on our disparity estimates, we bootstrapped each disparity measurement as described above for the stratigraphic time bins.
    \end{enumerate}

  \item Duration: time sub-samples defined by the duration of stratigraphic periods. 
  \begin{enumerate}[(i)]
    \item Time bins (equal). 
    We calculated disparity for the taxa in each time bin where time bin size was defined by the mean \textit{duration} of the stratigraphic period (age or epoch), and bootstrapped the disparity values as described above.
    \item Time-slices (equidistant).
    We calculated disparity using our time-slicing approach where the interval between slices, was defined by the mean \textit{duration} of the stratigraphic period (age or epoch).
    We used the six time-slicing methods and bootstrapped as described above.
  \end{enumerate}

  \item Number: time sub-samples defined by the number of stratigraphic periods. 
  \begin{enumerate}[(i)]
    \item Time bins (equal). 
    We calculated disparity for the taxa in each time bin where the number of time bins was defined by the \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest, and bootstrapped the disparity values as described above.
    \item Time-slices (equidistant).
    We calculated disparity using our time-slicing approach where the number of slices, was defined by the \textit{number} of stratigraphic periods (ages or epochs) in the time frame of interest.
    We used the six time-slicing methods and bootstrapped as described above.
  \end{enumerate}

@@

Time-slicing has several advantages over time binning (using either equally or unequally sized bins) approaches.
First, it allows us to use as much of the information available to us, in the form of phylogenetic relationships and ancestral taxa, as possible. 
This increases our ability to investigate key biological questions, such as how do various drivers influence morphological diversity through time, and how do mass extinctions influence disparity \citep{Brusatte12092008,Foote29111996,friedmanexplosive2010}, both by increasing the statistical power of analyses and through the availability of data at key time points in the history of our groups.
Second, we are able to be more explicit about the mode of evolution in our clades; in time-slicing we can apply punctuated or gradual models of trait change rather than making an assumption of punctuated evolution.
This may be important, as gradual change is a common pattern of trait evolution in the fossil record \citep{Hunt20112007}.

Of course the method also has limitations.
The main one of these is a practical one; it requires a time-calibrated phylogeny which are not available for all palaeontological datasets.
Furthermore, like most phylogeny based methods, time-slicing depends on ancestral state estimations.
Care should be taken in interpreting these, as they are highly dependent on the data and models used for the estimations \citep{Slateretal2012,doi:10.1080/10635150801910451}.
Finally, the difference between the time-binning and time-slicing methods could also simply be favoured towards time binning due to the nature of the fossil record itself.
The rates of sedimentation varies in time a space and can be really sporadic, influencing the groups found within and their temporal distribution (e.g. vertebrates, c.f. foraminifera).
In this case, different beds could represent different ``packages'' of fauna through time separated by gaps thus effectively binning the data rather than slicing.
For example, slicing through such strata will yield similar results whether where in time does the slice occur.
It is important to note however, that, as specified above, the time slicing method also includes ancestral estimations data (whether through the nodes or the branches) that are by definition not available in the fossil record and thus are not influenced by its nature.
